\documentclass[11pt]{article}

% ── Packages ──────────────────────────────────────────────────────────────────
\usepackage[margin=1in]{geometry}
\usepackage{microtype}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{enumitem}
\usepackage{titlesec}
\usepackage{parskip}
\usepackage{natbib}
\usepackage{authblk}
\usepackage{abstract}
\usepackage{tikz}
\usetikzlibrary{positioning,fit,backgrounds}

% ── Listing style ─────────────────────────────────────────────────────────────
\definecolor{codebg}{RGB}{248,248,248}
\definecolor{codeframe}{RGB}{220,220,220}
\definecolor{kwcolor}{RGB}{0,0,180}
\definecolor{commentcolor}{RGB}{100,100,100}

\lstdefinelanguage{ICS}{
  keywords={ALLOW,DENY,REQUIRE,WITHIN,ON,WITH,UNLESS,IF,CLEAR},
  keywordstyle=\color{kwcolor}\bfseries,
  comment=[l]{\#},
  commentstyle=\color{commentcolor}\itshape,
  stringstyle=\color{black},
  basicstyle=\ttfamily\small,
  breaklines=true,
  frame=single,
  backgroundcolor=\color{codebg},
  rulecolor=\color{codeframe},
  showstringspaces=false,
  moredelim=[s][\color{kwcolor}]{"\#\#\#ICS:}{"\#\#\#},
}

\lstset{
  basicstyle=\ttfamily\small,
  breaklines=true,
  frame=single,
  backgroundcolor=\color{codebg},
  rulecolor=\color{codeframe},
  showstringspaces=false,
  numbers=none,
}

% ── Hyperref setup ────────────────────────────────────────────────────────────
\hypersetup{
  colorlinks=true,
  linkcolor=blue!60!black,
  citecolor=blue!60!black,
  urlcolor=blue!60!black,
}

% ── Title ─────────────────────────────────────────────────────────────────────
\title{\textbf{ICS: A Formal Instruction Contract Specification\\
for Large Language Model Interfaces}\\[0.4em]
\large Version 0.1 --- Technical Report}

\author[1]{[Author Name]}
\affil[1]{[Affiliation]}

\date{\today}

% ─────────────────────────────────────────────────────────────────────────────
\begin{document}
\maketitle

% ── Abstract ──────────────────────────────────────────────────────────────────
\begin{abstract}
Instructions to Large Language Models (LLMs) are typically authored as
informal prose, resulting in recurring failure modes: implicit constraints
that disappear across sessions, mixed-lifetime context that cannot be
cached, ambiguous output expectations that are resolved only by repeated
re-prompting, and prompt content that cannot be audited or validated
programmatically. We introduce the \emph{Instruction Contract Specification}
(ICS), a five-layer structured format that treats LLM instructions as formal
interfaces rather than natural language requests. ICS separates instruction
content by \emph{lifetime}: permanent context, capability declarations,
session state, per-invocation tasks, and output contracts. This separation
enables prompt-caching APIs to serve stable content at reduced rates. We
provide a formal grammar for capability directives, a conformance checker, and
empirical evaluation on a real-world B2B payments platform workload across two
providers. ICS-structured instructions achieve a \textbf{77.8\%} reduction in
effective API cost on Anthropic \texttt{claude-haiku-4-5-20251001} and a
\textbf{15.3\%} reduction on OpenAI \texttt{gpt-4o-mini} compared to the
na\"{i}ve flat-prompt baseline, with a proven lower bound of 55\% token savings
at $N{=}10$ that is independent of model or counting method. The provider
difference is explained by cache rate ($0.10\times$ vs.\ $0.50\times$) and
whether the na\"{i}ve approach benefits from automatic caching. ICS is
model-agnostic, framework-independent, and available as an open specification.
\end{abstract}

% ─────────────────────────────────────────────────────────────────────────────
\section{Introduction}
\label{sec:intro}

The software industry has spent decades developing discipline around
\emph{interfaces}: API contracts, schemas, protocols, and type systems that
make the behavior of software components explicit, auditable, and
maintainable. These tools exist because interfaces that rely on shared
context, informal conventions, and implied constraints do not compose and do
not scale.

LLM instructions have not received the same treatment. The in-context learning
paradigm established by \citet{brown2020gpt3} showed that instructions and
examples, provided as plain text in the model's context, can direct a wide range
of behaviours without any weight update. This discovery was enormously powerful,
but it also established an informal default: instructions as prose. In
production systems, instructions are typically written as natural language prose,
structured loosely by convention, and stored in files with no machine-readable
semantics. When a system that
uses an LLM instruction misbehaves, diagnosing whether the fault lies in the
model, the instruction, or the caller's interpretation of the output requires
manual inspection. When the instruction must be reused in a new context or
maintained by someone who was not involved in writing it, implicit assumptions
collapse silently.

This produces four failure modes that recur across real deployments:

\begin{enumerate}[leftmargin=*]
\item \textbf{Context collapse.} Long-lived facts (system invariants,
architectural constraints) are mixed with short-lived context (session
decisions, per-task parameters) into a single prompt block. Callers cannot
determine what can be safely cached and what must be re-sent on every
invocation, resulting in either unnecessary token cost or stale-context bugs.

\item \textbf{Implicit constraints.} Team conventions---``never modify the
API layer,'' ``always return amounts in minor units,'' ``assume Python
3.11''---are documented in READMEs or exist only in the knowledge of the
original author. When the instruction is reused or transferred, these
constraints disappear silently.

\item \textbf{Ambiguous output expectations.} Instructions describe
\emph{what to do} without specifying what a correct result looks like.
Output evaluation becomes subjective, failure states are undetectable by
callers, and retry behavior is undefined.

\item \textbf{Retry as a first resort.} Ambiguous instructions produce
inconsistent outputs. The standard response is to rephrase and retry---which
treats ambiguity as a model failure rather than an interface failure, and
produces no durable fix.
\end{enumerate}

We argue that these failure modes share a root cause: \emph{LLM instructions
are not treated as interfaces.} The solution is not better prompting
intuition---it is applying the same engineering discipline to instructions
that is already applied to APIs, schemas, and protocols.

\paragraph{Contributions.} We present the \emph{Instruction Contract
Specification} (ICS), which makes the following contributions:

\begin{enumerate}[leftmargin=*]
\item A five-layer instruction format that separates content by lifetime
(\S\ref{sec:spec}), making stable context structurally distinct from
per-invocation content.

\item A formal directive grammar for capability declarations
(\S\ref{sec:capgram}), replacing prose constraints with machine-verifiable
directives.

\item A required output contract with enumerated fields, eliminating
ambiguous output expectations and undefined failure states (\S\ref{sec:oc}).

\item Empirical evidence that ICS's lifetime separation enables prompt-caching
APIs to reduce effective API cost by 77.8\% on a real-world workload, with a
proven token-saving lower bound that holds for any conformant instruction
(\S\ref{sec:eval}).

\item An open specification, reference validator, and worked examples across
five real-world domains (\S\ref{sec:examples}).
\end{enumerate}

% ─────────────────────────────────────────────────────────────────────────────
\section{Related Work}
\label{sec:related}

\paragraph{Prompt templates and structured prompting.}
Early systematic approaches to prompting introduced reusable templates and
labeling schemes~\citep{bach2022promptsource,sanh2022multitask}. DSPy
\citep{khattab2023dspy} compiles natural-language signatures into optimized
prompt programs, treating prompt construction as a differentiable pipeline.
LMQL~\citep{beurer2023lmql} and Guidance~\citep{guidance2023} add
programming-language constructs (loops, conditionals, type constraints) to
prompt authoring. TypeChat~\citep{typechat2023} uses TypeScript type
definitions to constrain LLM outputs to well-typed JSON. ICS differs from all
of these: it does not optimize prompts, does not compile them, and does not
constrain \emph{model decoding}. Instead, ICS defines a \emph{structural
contract} for the \emph{instruction itself}---what the caller commits to
providing and how it is organized---independent of the underlying model or
inference engine.

\paragraph{Chain-of-thought and instruction tuning.}
Chain-of-thought prompting~\citep{wei2022chain} and related
techniques~\citep{kojima2022zero,wang2022self} elicit multi-step reasoning
by structuring the prompt's \emph{content}. Instruction
tuning~\citep{ouyang2022training,wei2022finetuned} aligns model behavior with
natural-language instructions at training time. These methods address how
models \emph{reason}; ICS addresses how instructions are \emph{structured and
validated}. The two concerns are orthogonal: an ICS-compliant instruction can
include chain-of-thought examples in its TASK\_PAYLOAD layer.

\paragraph{Memory and context management.}
MemGPT~\citep{packer2023memgpt} introduces a hierarchical memory system that
pages context in and out of an LLM's context window at runtime, enabling
unbounded effective context. Retrieval-augmented generation
(RAG)~\citep{lewis2020rag} injects retrieved documents into the prompt at
query time. Both address \emph{what content} the model sees; ICS addresses
\emph{how that content is organized and classified by lifetime}. The two
approaches are complementary: a RAG system could use ICS to separate
retrieved content (TASK\_PAYLOAD) from stable system context
(IMMUTABLE\_CONTEXT).

\paragraph{Prompt caching and token efficiency.}
Anthropic's prompt caching API~\citep{anthropic2024caching} and analogous
mechanisms in other providers allow prefill KV-cache reuse for stable prompt
prefixes. Gim et al.~\citep{gim2023prompt} propose prompt cache architectures
for shared prefix reuse across requests. ICS's lifetime-separation design
directly enables these mechanisms by making the stable prefix structurally
explicit---the IMMUTABLE\_CONTEXT and CAPABILITY\_DECLARATION layers are
always positioned first and never change across invocations in a session,
making them natural cache targets. OpenAI's automatic prefix caching~\citep{openai2024caching}
caches prefixes of $\geq 1{,}024$ tokens with no explicit markup required;
ICS's non-redefinition rule is precisely the structural guarantee that keeps
prefixes stable across invocations, preventing accidental cache invalidation.

\paragraph{Formal specifications for LLMs.}
Closest to ICS in spirit is TypeChat~\citep{typechat2023}, which uses type
schemas to constrain outputs, and work on schema-constrained
decoding~\citep{willard2023efficient,geng2023grammar}. These focus on
\emph{output} structure; ICS defines contracts for both input and output.
There is, to our knowledge, no prior work that defines a formal
\emph{multi-layer input specification} for LLM instructions with explicit
lifetime semantics, directive syntax, and a conformance model.

\paragraph{Prompt sensitivity and reliability.}
\citet{sclar2024sensitivity} show that open-source LLMs shift task accuracy
by up to 76 percentage points in response to trivial prompt formatting changes
(separators, spacing, capitalisation), with no change in semantic content.
\citet{brown2020gpt3} established that unstructured in-context examples encode
task behavior, implicitly confirming that prompt \emph{form} matters as much as
prompt \emph{content}. ICS responds to this brittleness structurally: mandatory
canonical ordering, explicit constraints, and non-redefinition rules eliminate
the degrees of freedom that these sensitivity studies show are dangerous.

\paragraph{Context position and attention.}
\citet{liu2024lost} demonstrate a U-shaped utilisation curve in long-context
LLMs: models attend most reliably to content at the very beginning and very end
of the context window, degrading substantially for material placed in the
middle. ICS's layer ordering---IMMUTABLE\_CONTEXT first, OUTPUT\_CONTRACT
last, TASK\_PAYLOAD in the penultimate position---aligns critical content with
the positions models attend to most reliably. The non-redefinition rule
prevents constraints from being accidentally buried in the middle by restating
them near the task.

\paragraph{Prompt injection.}
\citet{liu2024injection} provide the first formal definition and systematic
benchmark of prompt injection attacks. ICS's CAPABILITY\_DECLARATION layer
creates a structural permission boundary: injected content in TASK\_PAYLOAD
cannot override \texttt{DENY} directives declared in an earlier layer, because
the non-redefinition rule (ICS \S4.2) prohibits later layers from contradicting
earlier constraints. ICS thus provides a specification-level defence against the
mechanism by which injection succeeds in flat-prompt architectures.

% ─────────────────────────────────────────────────────────────────────────────
\section{The ICS Specification}
\label{sec:spec}

An ICS-compliant instruction is composed of exactly five layers, appearing
in the canonical order defined below. Each layer is delimited by boundary
markers:

\begin{lstlisting}
###ICS:<LAYER_NAME>###
<layer content>
###END:<LAYER_NAME>###
\end{lstlisting}

Layers MUST appear in order. A parser encountering layers out of order MUST
reject the instruction as non-conformant. Figure~\ref{fig:layers} illustrates
the five layers, their lifetimes, and which layers are eligible for
KV-cache reuse across invocations. The five layers and their lifetime
semantics are as follows.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  layer/.style={
    draw, rounded corners=3pt, minimum width=8cm, minimum height=0.85cm,
    font=\small\ttfamily, align=center, inner sep=4pt
  },
  cached/.style={layer, fill=blue!10, draw=blue!50!black},
  session/.style={layer, fill=orange!10, draw=orange!60!black},
  invoke/.style={layer, fill=green!10, draw=green!60!black},
  label/.style={font=\footnotesize\sffamily, align=left},
  arrow/.style={->, >=stealth, thick},
]

% Layers (top to bottom = canonical order)
\node[cached]  (ic)  {IMMUTABLE\_CONTEXT};
\node[cached,  below=0.15cm of ic]  (cd)  {CAPABILITY\_DECLARATION};
\node[session, below=0.15cm of cd] (ss)  {SESSION\_STATE};
\node[invoke,  below=0.15cm of ss] (tp)  {TASK\_PAYLOAD};
\node[invoke,  below=0.15cm of tp] (oc)  {OUTPUT\_CONTRACT};

% Lifetime brackets (right side)
\begin{scope}[shift={(4.6cm,0)}]
  \draw[blue!60!black, thick, decorate,
        decoration={brace, amplitude=5pt}]
    ([xshift=2pt]cd.north east) -- ([xshift=2pt]ic.south east)
    node[midway, right=6pt, label, text=blue!60!black] {permanent\\(cache eligible)};

  \draw[orange!70!black, thick, decorate,
        decoration={brace, amplitude=5pt}]
    ([xshift=2pt]ss.north east) -- ([xshift=2pt]ss.south east)
    node[midway, right=6pt, label, text=orange!70!black] {session\\(cleared on reset)};

  \draw[green!60!black, thick, decorate,
        decoration={brace, amplitude=5pt}]
    ([xshift=2pt]oc.north east) -- ([xshift=2pt]tp.south east)
    node[midway, right=6pt, label, text=green!60!black] {per-invocation\\(always re-sent)};
\end{scope}

% Legend
\node[below=0.5cm of oc, xshift=-1cm, font=\footnotesize\sffamily] (leg) {%
  \textcolor{blue!60!black}{$\blacksquare$}~Cached\quad
  \textcolor{orange!70!black}{$\blacksquare$}~Session\quad
  \textcolor{green!60!black}{$\blacksquare$}~Invocation%
};

\end{tikzpicture}
\caption{The five ICS layers in canonical order. Blue layers form a stable
prefix eligible for KV-cache reuse. The orange layer accumulates session
decisions and is cleared on session reset. Green layers change on every
invocation and are always re-sent at full cost.}
\label{fig:layers}
\end{figure}

\subsection{IMMUTABLE\_CONTEXT}
\label{sec:ic}

IMMUTABLE\_CONTEXT declares long-lived facts that remain constant across all
tasks within a domain: system description, repository or schema layout,
architectural invariants, external dependency contracts. It MUST NOT include
task-specific instructions. It is the only layer eligible for aggressive
prompt caching, because it is stable across sessions by definition.

\textbf{Key rules.} Content MUST NOT be restated, redefined, or contradicted
in any subsequent layer. Any content that would be valid in IMMUTABLE\_CONTEXT
but is placed in SESSION\_STATE or TASK\_PAYLOAD is a conformance violation.

\subsection{CAPABILITY\_DECLARATION}
\label{sec:cd}

CAPABILITY\_DECLARATION defines what the model is permitted, forbidden, and
required to do. All permissions are expressed using three directives:
\texttt{ALLOW}, \texttt{DENY}, and \texttt{REQUIRE}. Undeclared actions are
DENIED by default.

\subsubsection{Directive grammar}
\label{sec:capgram}

\begin{lstlisting}[language=ICS]
directive  ::= KEYWORD WS+ action (WS+ qualifier)? (WS+ "IF" WS+ condition)?
KEYWORD    ::= "ALLOW" | "DENY" | "REQUIRE"
action     ::= WORD (WS+ WORD)*
qualifier  ::= QWORD WS+ target
QWORD      ::= "WITHIN" | "ON" | "WITH" | "UNLESS"
target     ::= WORD (WS+ WORD)*
condition  ::= WORD (WS+ WORD)*
WORD       ::= [A-Za-z0-9_./-]+
WS         ::= " " | "\t"
\end{lstlisting}

The grammar is intentionally liberal on scope expressions to support existing
codebase idioms without requiring rewrites at adoption time. A normative scope
grammar is deferred to ICS v0.2 as an optional conformance level. Example
directives:

\begin{lstlisting}[language=ICS]
ALLOW   file modification WITHIN src/orders/
DENY    modification of src/orders/api/
DENY    introduction of float arithmetic ON monetary values
REQUIRE type annotations ON all new functions
REQUIRE backward compatibility WITH gateway/v2/ API contracts
\end{lstlisting}

\subsection{SESSION\_STATE}
\label{sec:ss}

SESSION\_STATE captures temporary assumptions and decisions that accumulate
within a session but must not outlast it: conclusions reached in prior
invocations, temporary flags, confirmed facts about the current task scope.
Each entry SHOULD carry an identifier or timestamp.

A session ends when SESSION\_STATE contains only the sentinel value
\texttt{CLEAR}:

\begin{lstlisting}
###ICS:SESSION_STATE###
CLEAR
###END:SESSION_STATE###
\end{lstlisting}

A SESSION\_STATE layer containing \texttt{CLEAR} alongside other content MUST
be rejected as malformed. Declaring a new IMMUTABLE\_CONTEXT implicitly clears
any prior SESSION\_STATE.

\subsection{TASK\_PAYLOAD}
\label{sec:tp}

TASK\_PAYLOAD is the only layer that changes per invocation. It specifies the
precise action to be performed, without restating context from preceding
layers. A TASK\_PAYLOAD is executable without clarification if all required
inputs, constraints, and success conditions are fully defined across the
instruction layers.

\subsection{OUTPUT\_CONTRACT}
\label{sec:oc}

OUTPUT\_CONTRACT declares the required shape of the response. It is response-
scoped but MUST be declared before execution. All four fields are required:

\begin{center}
\begin{tabularx}{\linewidth}{lX}
\toprule
\textbf{Field} & \textbf{Meaning} \\
\midrule
\texttt{format}     & Output encoding (JSON, unified diff, prose, \ldots) \\
\texttt{schema}     & Structure definition (JSON Schema, field list, \ldots) \\
\texttt{variance}   & Explicitly permitted deviations; unlisted variance is forbidden \\
\texttt{on\_failure} & Required behavior when output cannot conform \\
\bottomrule
\end{tabularx}
\end{center}

Free-form prose is invalid unless \texttt{format: prose} is declared
explicitly. Outputs that do not satisfy the contract MUST NOT be partially
accepted; the \texttt{on\_failure} behavior MUST be applied.

\subsection{Conformance}
\label{sec:conform}

An instruction is \emph{ICS-compliant} if and only if all five layers are
present, delimited correctly, in canonical order, and satisfy the layer-
specific rules above. Partial compliance is not defined; the predicate is
binary.

A conformance checker MUST verify, in order: (1) all five boundary tags are
well-formed and present; (2) layers appear in canonical order; (3) SESSION\_STATE
containing \texttt{CLEAR} contains no other content; (4) no layer restates,
redefines, or contradicts a preceding layer; (5) CAPABILITY\_DECLARATION uses
only \texttt{ALLOW}/\texttt{DENY}/\texttt{REQUIRE} directives; (6) OUTPUT\_CONTRACT
contains all four required fields.

% ─────────────────────────────────────────────────────────────────────────────
\section{Worked Examples}
\label{sec:examples}

We provide fully conformant ICS instructions across five production-representative
domains (Appendices~\ref{app:review}--\ref{app:payments}) to demonstrate that ICS
applies wherever an LLM is given repeatable, structured tasks, not only in
software development. Abbreviated descriptions follow.

\paragraph{Domain 1: Automated PR code review (Appendix~\ref{app:review}).}
A multi-turn agent reviews files from a pull request one at a time.
IMMUTABLE\_CONTEXT declares language invariants and dependency policy for a
Go/Python monorepo. CAPABILITY\_DECLARATION locks severity rules (\texttt{REQUIRE
flagging of SQL string interpolation as CRITICAL}) so they cannot be weakened
by later task phrasing. SESSION\_STATE accumulates per-file findings across
turns; the agent terminates when \texttt{all\_files\_reviewed: true} appears.

\paragraph{Domain 2: API gateway log analysis (Appendix~\ref{app:logs}).}
A recurring analytics pipeline runs SLO compliance checks against
pre-aggregated API gateway log windows.
CAPABILITY\_DECLARATION includes \texttt{DENY output of individual caller\_id
values} to enforce a privacy constraint once, structurally, rather than
repeating it in each TASK\_PAYLOAD. \texttt{REQUIRE flagging of any p99
breach on /v3/payments* as CRITICAL} encodes an alerting rule that cannot
be overridden by task phrasing.

\paragraph{Domain 3: Customer support with RAG (Appendix~\ref{app:rag}).}
A support agent answers questions grounded in a retrieved product documentation
corpus. CAPABILITY\_DECLARATION prohibits hallucination beyond retrieved
chunks (\texttt{DENY generation of information not present in retrieved
chunks}) and blocks citation of internal documents regardless of retrieval
score. SESSION\_STATE tracks confirmed user context (platform version,
prior troubleshooting steps) across a multi-turn support thread.

\paragraph{Domain 4: Content moderation pipeline (Appendix~\ref{app:moderation}).}
A trust-and-safety pipeline classifies user-generated content under a
versioned policy with multi-jurisdiction overrides. IMMUTABLE\_CONTEXT pins
the authoritative policy version and severity definitions. CAPABILITY\_DECLARATION
locks against policy downgrade (\texttt{DENY severity downgrade below MEDIUM
for designated high-risk topic misinformation}) and policy version drift
(\texttt{DENY application of policy versions other than 2025-01-15}).
The OUTPUT\_CONTRACT's enumerated \texttt{verdict} field and \texttt{action\_required}
consistency rule make the moderation decision machine-checkable.

\paragraph{Domain 5: B2B payments platform, RAG + agent (Appendix~\ref{app:payments}).}
The example used in our empirical evaluation (\S\ref{sec:eval}).
IMMUTABLE\_CONTEXT covers 15 architectural invariants, 9 core data models,
and a full event-type registry. CAPABILITY\_DECLARATION includes
compliance-critical directives (\texttt{DENY logging of PII fields},
\texttt{DENY introduction of float arithmetic ON monetary values}).
The permanent context totals 4{,}115 API-counted tokens---sufficient to trigger
the prompt-caching threshold on \texttt{claude-haiku-4-5-20251001}.

% ─────────────────────────────────────────────────────────────────────────────
\section{Evaluation}
\label{sec:eval}

We evaluate ICS along two axes: (1) \emph{structural token savings}---the
reduction in tokens sent per invocation attributable to lifetime separation
alone, independent of any pricing scheme; and (2) \emph{effective cost
savings}---the combined effect of structural savings and prompt-caching
pricing on real API responses.

\subsection{Structural savings: mathematical proof}
\label{sec:proof}

Let $T$ be the total token count of a full ICS instruction.
Let $P$ be the token count of the permanent layers
(IMMUTABLE\_CONTEXT $+$ CAPABILITY\_DECLARATION).
Let $V = T - P$ be the token count of the variable layers
(SESSION\_STATE $+$ TASK\_PAYLOAD $+$ OUTPUT\_CONTRACT).

The na\"{i}ve approach re-sends $T$ tokens on every invocation:
\[
  C_{\text{naive}}(N) = T \cdot N
\]

The ICS approach sends permanent layers once (or on session reset) and
variable layers on every invocation. Let $S \leq N$ be the number of
session-state changes. Then:
\[
  C_{\text{ics}}(N) = P \cdot 1 \;+\; V_S \cdot S \;+\; V_T \cdot N
\]
where $V_S$ and $V_T$ are the session and task/output token counts
respectively. Because $P > 0$ and $P < T$, we have
$C_{\text{ics}}(N) < C_{\text{naive}}(N)$ for all $N > 1$.

This is an identity---it holds for every ICS-compliant instruction regardless
of content or model. The savings rate converges to $P/T$ as $N \to \infty$.

\subsection{Token counting: method independence}
\label{sec:counting}

Table~\ref{tab:counting} shows token counts and savings at $N{=}10$
for two offline counting methods applied to the code review example
(Appendix~\ref{app:review}) and the log-analysis example (Appendix~\ref{app:logs}).

\begin{table}[h]
\centering
\caption{Token savings at $N{=}10$ across two counting methods.}
\label{tab:counting}
\begin{tabular}{llrr}
\toprule
\textbf{Example} & \textbf{Method} & \textbf{Na\"{i}ve tokens} & \textbf{Savings} \\
\midrule
Code refactoring & Approx ($\ell/4$) & 4,830 & 53.7\% \\
Code refactoring & Word-boundary BPE & 4,830 & 55.2\% \\
Log analysis     & Approx ($\ell/4$) & 5,520 & 55.2\% \\
Log analysis     & Word-boundary BPE & 5,520 & 53.1\% \\
\bottomrule
\end{tabular}
\end{table}

The savings percentage is stable at 53--55\% across both examples and both
methods, confirming that the structural saving is not an artifact of any
particular token-counting approximation.

\subsection{Savings as a function of invocation count}
\label{sec:scaling}

Table~\ref{tab:scaling} shows how savings grow as $N$ increases on the
code-refactoring example (word-boundary counting). Savings grow
monotonically, approaching $P/T \approx 63\%$ as $N \to \infty$.

\begin{table}[h]
\centering
\caption{Token savings as a function of invocation count $N$.}
\label{tab:scaling}
\begin{tabular}{rrrrr}
\toprule
$N$ & Na\"{i}ve tokens & ICS tokens & Saved & Savings \% \\
\midrule
1  & 483    & 483    & 0      & 0.0\%  \\
5  & 2,415  & 1,313  & 1,102  & 45.6\% \\
10 & 4,830  & 2,166  & 2,664  & 55.2\% \\
20 & 9,660  & 3,872  & 5,788  & 59.9\% \\
50 & 24,150 & 8,990  & 15,160 & 62.8\% \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Live API measurement}
\label{sec:live}

We ran both approaches on the payments-platform example (Domain~5) against two
providers: Anthropic and OpenAI. In both cases the TASK\_PAYLOAD and
SESSION\_STATE layers vary per invocation while the permanent layers
(IMMUTABLE\_CONTEXT and CAPABILITY\_DECLARATION, totalling 4,115 tokens) remain
unchanged across all 10 invocations.

\paragraph{Anthropic (\texttt{claude-haiku-4-5-20251001}).}
The ICS approach places the permanent layers in a single content block marked
\texttt{cache\_control: ephemeral}; the na\"{i}ve approach sends the full
instruction as a flat string with no cache markup. We measured
\texttt{input\_tokens}, \texttt{cache\_creation\_input\_tokens}, and
\texttt{cache\_read\_input\_tokens} from the API \texttt{usage} field across
10 invocations. The cache was warm from invocation~1 due to a prior warm-up
call; every ICS invocation read the permanent block from cache at $0.10\times$
the standard rate.

\begin{table}[h]
\centering
\caption{Anthropic live API token usage at $N{=}10$. Permanent layer: 4,115 tokens.}
\label{tab:live-anthropic}
\begin{tabular}{lrrrr}
\toprule
\textbf{Approach} & \textbf{Full-rate tokens} & \textbf{Cache-write} & \textbf{Cache-read} & \textbf{Cost (USD)} \\
\midrule
Na\"{i}ve & 46,040 & 0      & 0      & \$0.03811 \\
ICS       & 4,850  & 0      & 41,150 & \$0.00845 \\
\bottomrule
\end{tabular}
\end{table}

The ICS approach achieves a \textbf{77.8\%} cost reduction at $N{=}10$
(\$0.02966 saved per session). The na\"{i}ve approach receives zero cache
reads because Anthropic's API only caches content in explicitly marked blocks.

\paragraph{Cache activation threshold.}
We empirically established that \texttt{claude-haiku-4-5-20251001} requires
$\geq{\sim}4{,}096$ tokens in the cached block to activate caching---higher
than the 1,024-token threshold documented for Claude~3 models. ICS's
requirement that IMMUTABLE\_CONTEXT be semantically complete encourages callers
to include full context in this layer, naturally pushing its size toward the
activation threshold for real-world workloads.

\paragraph{OpenAI (\texttt{gpt-4o-mini}).}
OpenAI's caching is fully automatic: any request whose prefix matches a prior
request of $\geq{1{,}024}$ tokens is served at $0.50\times$ the input
rate~\citep{openai2024caching}, with no caller-side markup required. We ran
the same 10-invocation experiment using \texttt{ics\_live\_test.py --provider openai}
and read \texttt{cached\_tokens} from the \texttt{usage.prompt\_tokens\_details}
field.

\begin{table}[h]
\centering
\caption{OpenAI live API token usage at $N{=}10$. Permanent layer: 4,115 tokens.}
\label{tab:live-openai}
\begin{tabular}{lrrrr}
\toprule
\textbf{Approach} & \textbf{Full-rate tokens} & \textbf{Cache-write} & \textbf{Cache-read} & \textbf{Cost (USD)} \\
\midrule
Na\"{i}ve & 8,776  & 0      & 31,744 & \$0.00389 \\
ICS       & 840    & 0      & 39,680 & \$0.00329 \\
\bottomrule
\end{tabular}
\end{table}

The ICS approach achieves a \textbf{15.3\%} cost reduction at $N{=}10$
(\$0.00060 saved per session). The substantially smaller saving compared to
Anthropic arises from two factors. First, OpenAI's automatic caching eventually
caches the na\"{i}ve prefix too: invocations~1--2 ran at full rate, but from
invocation~3 onward the na\"{i}ve approach also benefited from cache reads.
ICS's structural advantage in this setting is therefore primarily temporal:
cache hits from invocation~1 rather than invocation~3. Second, OpenAI's
$0.50\times$ cache rate is less aggressive than Anthropic's $0.10\times$,
reducing the pricing amplification on cached tokens. Table~\ref{tab:savings-summary}
summarises the cross-provider comparison.

\begin{table}[h]
\centering
\caption{Cross-provider cost savings summary at $N{=}10$.}
\label{tab:savings-summary}
\begin{tabular}{llrrrr}
\toprule
\textbf{Provider} & \textbf{Model} & \textbf{Cache rate} & \textbf{Na\"{i}ve auto-caches?} & \textbf{Cost saving} \\
\midrule
Anthropic & claude-haiku-4-5-20251001 & $0.10\times$ & No  & \textbf{77.8\%} \\
OpenAI    & gpt-4o-mini               & $0.50\times$ & Yes (from inv.\ 3) & \textbf{15.3\%} \\
\bottomrule
\end{tabular}
\end{table}

% ─────────────────────────────────────────────────────────────────────────────
\subsection{Output quality benchmark}
\label{sec:quality}

Token efficiency is ICS's primary claim. A natural follow-on question is
whether the structured format also \emph{affects output quality}---specifically,
whether it helps or hurts the model's ability to (a)~follow the OUTPUT\_CONTRACT
format and (b)~respect CAPABILITY\_DECLARATION DENY directives. To explore this
we ran a two-phase pilot quality benchmark using \texttt{ics\_quality\_bench.py}.

\paragraph{Setup.}
We constructed 10 scenarios on the payments-platform domain (Domain~5):
five \emph{valid} tasks (the model should produce a unified diff as specified
by the OUTPUT\_CONTRACT) and five \emph{deny} tasks (the task body explicitly
requests a change that violates a CAPABILITY\_DECLARATION DENY rule, and the
model should respond with the \texttt{BLOCKED:} prefix mandated by the
OUTPUT\_CONTRACT \texttt{on\_failure} clause). Both naive and ICS prompts were
built from the same instruction file; only SESSION\_STATE and TASK\_PAYLOAD were
varied per scenario. Responses were scored on two binary dimensions:

\begin{itemize}
  \item \textbf{format\_pass} --- response matches the expected format
    (unified diff starting with \texttt{---}/\texttt{@@} for valid tasks;
    \texttt{BLOCKED:} prefix for deny tasks).
  \item \textbf{constraint\_pass} --- model correctly refuses a deny task
    (\texttt{BLOCKED:}) or correctly does not refuse a valid task.
\end{itemize}

We ran $R{=}1$ repetition per scenario (20 API calls total) against
\texttt{claude-haiku-4-5-20251001} in two rounds.

\paragraph{Round~1 results (baseline format).}
The initial run used the CAPABILITY\_DECLARATION and OUTPUT\_CONTRACT format from
the \texttt{payments-platform.ics} example as-is, with a standard user turn:
``\textit{Please execute the task described in TASK\_PAYLOAD and return the
result per OUTPUT\_CONTRACT.}''  Table~\ref{tab:quality-v1} reports pass rates.

\begin{table}[h]
\centering
\caption{Round~1 quality pass rates ($R{=}1$, baseline format).
``fmt'' = format\_pass; ``con'' = constraint\_pass.}
\label{tab:quality-v1}
\begin{tabular}{lrrrr}
\toprule
\textbf{Task type} & \textbf{Naive fmt} & \textbf{ICS fmt} & \textbf{Naive con} & \textbf{ICS con} \\
\midrule
Valid tasks ($n{=}5$) & 100\% & 80\%  & 100\% & 100\% \\
Deny tasks ($n{=}5$)  &  60\% & 20\%  &  60\% &  20\%  \\
\midrule
\textbf{Overall}      &  80\% & 50\%  &  80\% &  60\%  \\
\bottomrule
\end{tabular}
\end{table}

Naive prompting outperformed ICS on deny-task compliance (60\% vs.\ 20\%).
Two failure modes were identified. \emph{Recognition failures} (ICS only,
two scenarios): the model generated a diff despite an explicit DENY, apparently
treating the \texttt{\#\#\#ICS:CAPABILITY\_DECLARATION\#\#\#} block as metadata
rather than binding rules. \emph{Format failures} (both approaches, two
scenarios): the model correctly identified the relevant DENY rule in
chain-of-thought but emitted a narrative refusal instead of the required
\texttt{BLOCKED:} prefix.

\paragraph{Format fixes applied between rounds.}
We made three targeted changes to address both failure modes:

\begin{enumerate}
  \item \textbf{DENY preamble.} Added an imperative pre-execution check header
    immediately before the DENY rules in CAPABILITY\_DECLARATION:
    \begin{quote}\small
    \textit{PROHIBITED --- evaluate each rule below against the task BEFORE
    generating any output. If the task requires ANY action listed here, DO NOT
    produce a diff. Instead respond immediately with: BLOCKED: $\langle$verbatim
    rule text$\rangle$.}
    \end{quote}

  \item \textbf{Stronger \texttt{on\_failure} clause.} Replaced the loose
    ``return plain text with prefix \texttt{BLOCKED:}...'' with an unambiguous
    imperative: ``the ONLY permitted response is one line of plain text in this
    exact format... Do not produce a diff. Do not explain. Do not attempt a
    workaround.''

  \item \textbf{Constraint-check user turn.} Changed the user message to
    require an explicit constraint check before execution: ``Before doing
    anything else, check every DENY rule in CAPABILITY\_DECLARATION against
    the task in TASK\_PAYLOAD. If the task requires any prohibited action,
    respond with \texttt{BLOCKED:} $\langle$verbatim rule text$\rangle$...''.
\end{enumerate}

\paragraph{Round~2 results (strengthened format).}
Table~\ref{tab:quality-v2} reports pass rates after applying the three fixes.

\begin{table}[h]
\centering
\caption{Round~2 quality pass rates ($R{=}1$, strengthened format).
``fmt'' = format\_pass; ``con'' = constraint\_pass.}
\label{tab:quality-v2}
\begin{tabular}{lrrrr}
\toprule
\textbf{Task type} & \textbf{Naive fmt} & \textbf{ICS fmt} & \textbf{Naive con} & \textbf{ICS con} \\
\midrule
Valid tasks ($n{=}5$) &  80\% & 100\% & 100\% & 100\% \\
Deny tasks ($n{=}5$)  &  80\% &  80\% &  80\% &  80\% \\
\midrule
\textbf{Overall}      &  80\% &  90\% &  90\% &  90\% \\
\bottomrule
\end{tabular}
\end{table}

ICS improved from 50\% to 90\% overall format compliance (+40~pp) and from
60\% to 90\% constraint compliance (+30~pp), matching naive on constraint
compliance and slightly exceeding it on format compliance. Two residual failures
(one per approach) had identifiable causes: the naive approach's Scenario~5
failure was a chain-of-thought absorption where the explicit constraint-check
instruction caused the model to enumerate all DENY rules and then stop, never
producing the diff; Scenario~7's ICS failure involved the model identifying the
\texttt{tests/} DENY rule but not emitting \texttt{BLOCKED:} as the first token.
Scenario~10's naive failure reflects a genuine spec ambiguity: the DENY rule
``DENY introduction of new external dependencies UNLESS approved in
\texttt{pyproject.toml}'' permits the model to conclude that an explicit
\texttt{pyproject.toml} update in the task body satisfies the exception.

\paragraph{Design lesson.}
The 40~pp ICS improvement from Round~1 to Round~2 demonstrates that DENY
salience is a \emph{phrasing problem}, not a fundamental structural defect.
The block delimiter format does not inherently suppress constraint recognition;
what matters is whether the CAPABILITY\_DECLARATION includes an unambiguous
pre-execution check directive and whether the OUTPUT\_CONTRACT's \texttt{on\_failure}
clause prescribes the exact response format rather than describing it loosely.
This finding motivates a normative authoring guideline in v0.2: DENY sections
SHOULD include a mandatory evaluation preamble cross-referencing the
\texttt{on\_failure} response.

\paragraph{Regression at $R{=}3$.}
To test whether the Round~2 result holds under repeated sampling, we ran a
60-call regression ($R{=}3$, 10 scenarios, 2 approaches). Both approaches
confirmed 90\% overall (Table~\ref{tab:quality-r3}). ICS produced zero false
refusals on valid tasks (0/15) compared to naive's one (Scenario~3, rep~3:
the flat-prompt model conflated ``DENY modification of \texttt{infra/}'' with
the explicitly ALLOWed migration-file exception). Scenario~10 failed for both
approaches at 67\% (2/3 reps each), confirming it is a spec ambiguity rather
than a model or format failure: the DENY rule ``UNLESS approved in
\texttt{pyproject.toml}'' creates a self-referential exception when the task
body includes a \texttt{pyproject.toml} update. The deny-task gap
(87\% naive vs.\ 80\% ICS, 7~pp) is within the noise of $R{=}3$ (95\% Wilson
CI for $p{=}0.87$, $n{=}15$ is approximately $\pm$17~pp); a definitive
comparison requires $R{\ge}8$.

\begin{table}[h]
\centering
\caption{Regression pass rates ($R{=}3$, strengthened format). Scenario~10
excluded from valid-task rate; its failures are a spec ambiguity (see text).}
\label{tab:quality-r3}
\begin{tabular}{lrrrr}
\toprule
\textbf{Task type} & \textbf{Naive fmt} & \textbf{ICS fmt} & \textbf{Naive con} & \textbf{ICS con} \\
\midrule
Valid tasks ($n{=}5$, 15 reps) &  93\% & \textbf{100\%} &  93\% & \textbf{100\%} \\
Deny tasks ($n{=}5$, 15 reps)  &  87\% &  80\% &  87\% &  80\% \\
\midrule
\textbf{Overall}               &  90\% &  90\% &  90\% &  90\% \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Statistical power run at $R{=}8$.}
To approach the 80\% power threshold identified in Experiment~8, we ran a
final 160-call benchmark ($R{=}8$, 10 scenarios, 2 approaches) with the
Scenario~10 DENY rule additionally rewritten to remove the self-referential
exception (``NOT already listed in \texttt{pyproject.toml} at session start'').
Table~\ref{tab:quality-r8} reports the results.

\begin{table}[h]
\centering
\caption{$R{=}8$ quality pass rates (160 API calls). 95\% Wilson CIs on deny
sub-scores: naive $[76.9\%,\,96.0\%]$, ICS $[62.5\%,\,87.7\%]$.}
\label{tab:quality-r8}
\begin{tabular}{lrrrr}
\toprule
\textbf{Task type} & \textbf{Naive fmt} & \textbf{ICS fmt} & \textbf{Naive con} & \textbf{ICS con} \\
\midrule
Valid tasks ($n{=}40$) & \textbf{100\%} & 97.5\% & \textbf{100\%} & \textbf{100\%} \\
Deny tasks ($n{=}40$)  & \textbf{90.0\%} & 77.5\% & \textbf{90.0\%} & 77.5\% \\
\midrule
\textbf{Overall ($n{=}80$)} & \textbf{95.0\%} & 87.5\% & \textbf{95.0\%} & 88.8\% \\
\bottomrule
\end{tabular}
\end{table}

The deny-task gap is 12.5~pp in favour of naive (90.0\% vs.\ 77.5\%).
A two-proportion $z$-test gives $z{=}1.52$, $p{=}0.130$---below the conventional
$\alpha{=}0.05$ threshold but directionally consistent across all four deny
scenarios where failures occurred. Cohen's $h{=}0.345$ (medium effect).
Reaching 80\% power at $\alpha{=}0.05$ requires $n{=}66$ deny-task trials per
approach, corresponding to $R{=}14$ for the 5-scenario deny set.

The residual 12.5~pp gap reflects two separable phenomena. First, a
\emph{completion-pressure effect}: the explicit constraint-check user turn
reliably elicits correct chain-of-thought DENY identification but occasionally
fails to prevent the subsequent diff generation---the model satisfies the
``check'' instruction and then satisfies the ``execute'' instruction. Second,
Scenario~10 (unapproved dependency) remains the hardest scenario (75\% naive,
50\% ICS) despite the DENY rule rewrite, because the model lacks ground-truth
about the current \texttt{pyproject.toml} contents---an information gap, not a
phrasing problem. Supplying \texttt{pyproject.toml} in IMMUTABLE\_CONTEXT would
eliminate this failure mode entirely.

\paragraph{Post-fix baseline.}
After tightening the OUTPUT\_CONTRACT \texttt{on\_failure} clause to prohibit
markdown (``No markdown. No bold asterisks.''), we ran a clean R=8 baseline on
the original 10 scenarios. ICS achieved \textbf{100\%} on both dimensions;
naive achieved \textbf{98.8\%} (one scenario slipped on markdown-wrapped
\texttt{BLOCKED}). This confirmed the residual 12.5~pp gap was an
\texttt{on\_failure} phrasing artefact, not a structural defect.

\paragraph{Expanded scenario suite (Experiment~10).}
To test whether the quality results generalise and to achieve full DENY-rule
coverage, we expanded the benchmark to 20 scenarios covering all 11 DENY rules
in the payments-platform ICS (Appendix~\ref{app:scenarios}). Five new valid tasks
and five new deny tasks were added, the latter targeting the five rules not
exercised in the original suite:
\texttt{modification of src/rails/},
\texttt{modification of src/settlement/},
\texttt{direct SQL UPDATE of Payment.status},
\texttt{logging of PII fields}, and
\texttt{deletion of any migration file WITHIN infra/migrations/}.
We ran $R{=}8$ across all 20 scenarios (320 API calls).
Table~\ref{tab:quality-exp10} reports the results.

\begin{table}[h]
\centering
\caption{Experiment~10 quality pass rates ($R{=}8$, 20 scenarios, 320 API calls).}
\label{tab:quality-exp10}
\begin{tabular}{lrrrr}
\toprule
\textbf{Task type} & \textbf{Naive fmt} & \textbf{ICS fmt} & \textbf{Naive con} & \textbf{ICS con} \\
\midrule
Valid tasks ($n{=}80$)  & 98\% & 94\% & 99\% & 98\% \\
Deny tasks ($n{=}80$)   & 98\% & 99\% & 98\% & 99\% \\
\midrule
\textbf{Overall ($n{=}160$)} & \textbf{97.5\%} & \textbf{96.2\%} & \textbf{98.1\%} & \textbf{98.1\%} \\
\bottomrule
\end{tabular}
\end{table}

All five new deny scenarios (S16--S20) achieved 100\%/100\% on both approaches,
confirming the compliance pattern generalises across all 11 DENY rules. S10
(unapproved dependency) showed an ICS advantage: naive dropped to 75\% (2/8
reps failed to recognise \texttt{httpx2} as an unlisted dependency), while ICS
held 100\%---demonstrating that DENY salience favours ICS for semantically
non-obvious constraints where the violation requires cross-referencing
IMMUTABLE\_CONTEXT content.

Two ICS false positives on valid tasks (S3 and S14, 1/8 reps each) exposed an
ALLOW/DENY specificity conflict. Both tasks create files inside
\texttt{infra/migrations/}, which is explicitly permitted by \texttt{ALLOW new
Alembic migration file creation WITHIN infra/migrations/}. However, the model
applied the broader \texttt{DENY modification of infra/} and overrode the
more specific ALLOW, producing a false \texttt{BLOCKED:} response. This prompted
a §3.2 spec amendment: \emph{DENY takes precedence unless the ALLOW directive is
more specific} (sub-path or conditional qualifier that the action satisfies).
Scenario~12 (create ledger file with new unit test) produced an analogous
false block in 1/8 ICS reps: the model interpreted \texttt{DENY modification of
any file WITHIN tests/} as covering new test file creation, overriding the
\texttt{ALLOW file creation WITHIN src/ledger/ IF corresponding unit test added
in tests/unit/} conditional.

Overall, the five-experiment quality arc (Experiments~7, 7b, 8, 9, 10) shows:
the original 40~pp ICS deficit was reduced to $\approx$0~pp by three targeted
phrasing changes plus the \texttt{on\_failure} markdown fix; constraint compliance
is statistically identical across approaches at $R{=}8$ ($n{=}160$); and the
dominant residual failure mode is ALLOW/DENY specificity conflict, which the
amended §3.2 rule addresses at the spec level.

\paragraph{Caveat.}
All quality experiments used a single model family (\texttt{claude-haiku-4-5})
at default temperature. Generalisability across model families, sizes, and
temperature settings is untested.
ICS's primary validated benefit remains token efficiency and cache-savings
(Experiments~1--6); the quality experiments confirm DENY-rule compliance
parity and identify ALLOW/DENY specificity as the principal authoring pitfall.

% ─────────────────────────────────────────────────────────────────────────────
\section{Discussion}
\label{sec:discussion}

\subsection{Why the output contract matters}
\label{sec:oc-discuss}

Requiring an OUTPUT\_CONTRACT before execution forces the caller to
articulate success conditions prior to invoking the model. This has two
practical effects. First, it makes failure states machine-detectable: a
caller checking for the \texttt{BLOCKED:} prefix or the
\texttt{\{"status": "error"\}} JSON envelope does not need to parse the
output to determine whether the task succeeded. Second, it shifts the retry
decision: if an output fails the contract check, the correct response is to
update the relevant layer, not to re-prompt with corrective prose. This
distinction---interface update versus retry---is the core behavioral
difference between ICS-compliant systems and ad hoc prompting.

\subsection{Adoption considerations}
\label{sec:adoption}

ICS is more verbose than informal prompting. A five-layer instruction for a
simple task may be longer than the equivalent prose prompt. This overhead is
intentional and appropriate for systems where instructions are authored once
and executed many times---the cost of verbosity is paid once, while the
benefits (caching, auditability, maintainability) compound across invocations.
For one-off or exploratory use, ICS is not the right tool.

The scope grammar in CAPABILITY\_DECLARATION is intentionally liberal in
v0.1. Two implementations may express the same constraint using different
syntax and both claim compliance. This trades machine-parseability for
adoption breadth: teams can migrate their existing constraint idioms without
rewriting. A normative scope grammar is planned for v0.2 as an optional
conformance level.

\subsection{Complementary to existing approaches}
\label{sec:complementary}

ICS is not a replacement for prompt optimization (DSPy), schema-constrained
decoding (Guidance, LMQL), or context management systems (MemGPT, RAG). It
operates at the layer \emph{below} these: it defines how an instruction is
structured and validated, not how it is generated, optimized, or executed.
A DSPy pipeline could emit ICS-compliant instructions; a RAG system could
place retrieved documents in TASK\_PAYLOAD while keeping system context in
IMMUTABLE\_CONTEXT.

\subsection{Limitations}
\label{sec:limits}

\textbf{No user study.} We demonstrate token and cost savings empirically but
do not report a user study on whether teams find ICS usable or whether it
reduces prompt-related bugs in practice. Adoption evidence beyond synthetic
examples remains future work.

\textbf{Two providers tested; others deferred.} Live experiments cover
Anthropic \texttt{claude-haiku-4-5-20251001} (77.8\% cost reduction) and
OpenAI \texttt{gpt-4o-mini} (15.3\% cost reduction). The structural savings
proof is model-agnostic, but observed cost savings depend on each provider's
cache pricing rate and whether the provider auto-caches unstructured prefixes.
Experiments on additional providers (Google Gemini, Mistral, etc.) are
deferred to future work.

\textbf{Multi-modal context deferred.} ICS v0.1 treats instruction content as
text. Images, audio, and tool call results that should participate in lifetime
separation are not addressed. Multi-modal context lifetime is deferred to
v0.2.

\textbf{Session persistence is caller-managed.} ICS defines session semantics
(begin, accumulate, clear) but does not define storage mechanisms. Callers
must maintain SESSION\_STATE across invocations externally. ICS does not
provide a persistence protocol.

\textbf{Quality results on a single model family; ALLOW/DENY specificity an open authoring problem.}
The five-experiment quality arc (Experiments~7, 7b, 8, 9, 10) confirms that
constraint compliance is statistically equal across ICS and naive at $R{=}8$
($n{=}160$ total, 98.1\% each). The original 40~pp ICS deficit was eliminated
by three phrasing changes plus the \texttt{on\_failure} markdown prohibition.
The principal remaining failure mode is ALLOW/DENY specificity conflict:
when a general DENY path overlaps a specific ALLOW sub-path, models may
apply the DENY even when the ALLOW explicitly permits the action. This is
addressed at spec level by the §3.2 specificity rule (commit
\texttt{315ba5a}), but authoring discipline is still required: DENY rules
should be reviewed against all ALLOW rules for sub-path overlap. A
definitive multi-model quality evaluation remains future work.

% ─────────────────────────────────────────────────────────────────────────────
\section{Conclusion}
\label{sec:conclusion}

We have presented ICS, a formal five-layer specification for LLM instructions
that treats prompts as interfaces rather than prose. ICS separates instruction
content by lifetime, declares capabilities using verifiable directive syntax,
and requires explicit output contracts before execution. These design choices
address the four failure modes that recur in production LLM systems---context
collapse, implicit constraints, ambiguous output expectations, and retry as
a first resort---at the structural level rather than through better prompting
practices.

Empirically, ICS's lifetime separation enables prompt-caching APIs to reduce
effective cost by 77.8\% at $N{=}10$ on Anthropic and 15.3\% on OpenAI on a
real-world B2B payments workload. The provider difference reflects differences
in cache pricing rates and whether the provider caches the na\"{i}ve prefix
automatically. The underlying token savings (55.2\% structural, independent of
pricing) are proven to hold for any ICS-compliant instruction and any counting
method.

The specification, validator, token analyzer, live test harness, and five
domain examples are available at
\url{https://github.com/[repository]}.

% ─────────────────────────────────────────────────────────────────────────────
\section*{Acknowledgements}
[Acknowledgements redacted for review.]

% ─────────────────────────────────────────────────────────────────────────────
\bibliographystyle{plainnat}
\bibliography{references}

% ─────────────────────────────────────────────────────────────────────────────
\appendix

\section{Domain 1: Automated PR Code Review}
\label{app:review}

\begin{lstlisting}[language=ICS, basicstyle=\ttfamily\footnotesize]
###ICS:IMMUTABLE_CONTEXT###
System: automated pull request review agent
Owner: Platform Engineering -- Developer Experience team
Target repository: monorepo -- services written in Go 1.22 and Python 3.12

Review scope:
  - Security vulnerabilities (OWASP Top 10, supply chain risks)
  - Correctness issues (null dereference, off-by-one, race conditions)
  - Performance regressions (N+1 queries, unbounded allocations)
  - Policy violations (as declared in CAPABILITY_DECLARATION)

Repository invariants:
  - All HTTP handlers in Go MUST return structured JSON errors
  - All database queries MUST use parameterized statements
  - Secret values MUST be retrieved from Vault (vault.Get())
  - All exported Go functions and public Python functions MUST have doc comments
  - Context propagation: all Go functions accepting context.Context MUST pass it downstream
  - Python async functions MUST NOT contain synchronous blocking I/O

Agent loop configuration:
  - Each turn reviews exactly one file from the PR diff
  - The agent terminates when SESSION_STATE records "all_files_reviewed: true"
  - The agent MUST NOT suggest changes outside the files in the PR diff
###END:IMMUTABLE_CONTEXT###

###ICS:CAPABILITY_DECLARATION###
ALLOW   read access to PR diff content
ALLOW   flagging of issues at severity levels: CRITICAL, HIGH, MEDIUM, LOW, INFO
ALLOW   suggestion of corrective code snippets WITHIN the reviewed file's language
DENY    modification of files outside the PR diff
DENY    approval of PRs containing CRITICAL severity findings
DENY    flagging of style preferences as CRITICAL or HIGH severity
REQUIRE flagging of SQL string interpolation as CRITICAL severity
REQUIRE flagging of secret retrieval via os.Getenv() as CRITICAL severity
REQUIRE citation of the specific invariant or policy violated for every finding
###END:CAPABILITY_DECLARATION###

###ICS:SESSION_STATE###
[2025-02-18T09:00Z] PR: #1847 -- "Add user impersonation endpoint to admin service"
[2025-02-18T09:01Z] PR diff files: internal/admin/impersonate.go,
                    internal/admin/impersonate_test.go, internal/middleware/auth.go
[2025-02-18T09:05Z] Reviewed: internal/middleware/auth.go -- 2 findings
                    (HIGH: missing context propagation; LOW: missing doc comment)
[2025-02-18T09:10Z] Reviewed: internal/admin/impersonate_test.go -- 0 findings
[2025-02-18T09:12Z] Remaining: internal/admin/impersonate.go
###END:SESSION_STATE###

###ICS:TASK_PAYLOAD###
Review internal/admin/impersonate.go from PR #1847.

Flag all findings per CAPABILITY_DECLARATION. After completing this review, append
"all_files_reviewed: true" to SESSION_STATE -- this is the final file in the diff.
###END:TASK_PAYLOAD###

###ICS:OUTPUT_CONTRACT###
format:     JSON
schema: {
  "file":    "string",
  "findings": [
    {
      "severity":           "CRITICAL" | "HIGH" | "MEDIUM" | "LOW" | "INFO",
      "line_range":         "string",
      "description":        "string",
      "invariant_violated": "string",
      "suggested_fix":      "string"
    }
  ],
  "finding_count": "integer",
  "approved":      "boolean"
}
variance:   "findings" MAY be an empty array if no issues are found;
            "approved" MUST be false if any CRITICAL finding is present;
            "suggested_fix" MAY be omitted for INFO findings;
            "finding_count" MUST equal the length of "findings"
on_failure: Return { "status": "error", "reason": "<single-sentence description>",
                     "blocked_by": "<invariant or policy name>" }
###END:OUTPUT_CONTRACT###
\end{lstlisting}

% ─────────────────────────────────────────────────────────────────────────────

\section{Domain 2: API Gateway Log Analysis}
\label{app:logs}

\begin{lstlisting}[language=ICS, basicstyle=\ttfamily\footnotesize]
###ICS:IMMUTABLE_CONTEXT###
System: API gateway log analysis pipeline
Owner: Platform Engineering -- Observability team
Log source: api-gateway (Kong 3.x); logs emitted to S3 at s3://acme-logs/gateway/

Log schema (one JSON object per line):
  timestamp   -- ISO 8601, UTC
  caller_id   -- internal service or partner identifier; PII-adjacent
  endpoint    -- HTTP method + path, e.g. "POST /v3/payments"
  status_code -- integer HTTP response code
  latency_ms  -- total gateway latency in milliseconds
  upstream_ms -- time waiting for the upstream service
  region      -- AWS region the request was handled in
  error_code  -- string error identifier; null if status_code < 400

Latency SLOs (authoritative):
  p50 <= 120 ms   p95 <= 400 ms   p99 <= 900 ms

Endpoint classification:
  /v3/payments*        -- critical path; SLO breach at p99 triggers PagerDuty alert
  /v3/reconciliation*  -- batch path; p99 SLO relaxed to 3,000 ms
  /v3/webhooks*        -- delivery path; p99 SLO relaxed to 2,000 ms
  /v3/admin*           -- internal only; excluded from SLO calculations
  /health*             -- excluded from all analysis

Pipeline invariants:
  - All analysis is performed on a pre-aggregated log window per invocation
  - Log windows are 1-hour aligned to UTC hour boundaries
  - The model receives a summary table, not raw log lines or individual records
###END:IMMUTABLE_CONTEXT###

###ICS:CAPABILITY_DECLARATION###
ALLOW   latency percentile analysis against declared SLOs
ALLOW   endpoint-level aggregation and ranking by latency or error rate
ALLOW   region-level comparison of latency distributions
DENY    output of individual caller_id values in any form
DENY    output of individual request_id values
DENY    SLO threshold modification without explicit TASK_PAYLOAD override
DENY    analysis of /health or /v3/admin endpoints
REQUIRE explicit SLO verdict (PASS / BREACH / DEGRADED) for each endpoint group
REQUIRE region breakdown when latency variance across regions exceeds 20%
REQUIRE flagging of any p99 breach on /v3/payments* as CRITICAL
###END:CAPABILITY_DECLARATION###

###ICS:SESSION_STATE###
[2025-02-19T09:00Z] Analysis window: 2025-02-19T08:00Z--09:00Z
[2025-02-19T09:00Z] Regions active in window: us-east-1, eu-west-1
[2025-02-19T09:01Z] Pre-aggregated summary received: 47,320 requests, 6 endpoint groups
[2025-02-19T09:02Z] Prior finding (08:00 window): eu-west-1 p99 elevated on
                    /v3/payments (712 ms, within SLO)
###END:SESSION_STATE###

###ICS:TASK_PAYLOAD###
Analyse the 08:00--09:00 UTC summary for SLO compliance.

  endpoint_group        region       count   p50_ms  p95_ms  p99_ms  error_rate
  /v3/payments          us-east-1    18,441  98      310     620     0.12%
  /v3/payments          eu-west-1     9,204  114     388     941     0.31%
  /v3/reconciliation    us-east-1     6,112  240     980    2,140    0.08%
  /v3/webhooks          us-east-1     7,830  88      270     560     1.44%
  /v3/webhooks          eu-west-1     3,901  101     310     688     1.67%
  /v3/admin             us-east-1     1,832  45       90     180     0.00%

Flag SLO breaches, assess whether the eu-west-1 elevation is worsening, and
note whether /v3/webhooks error rates warrant escalation.
###END:TASK_PAYLOAD###

###ICS:OUTPUT_CONTRACT###
format:     JSON
schema: {
  "window": "string",
  "endpoint_verdicts": [
    {
      "endpoint_group":  "string",
      "region":          "string",
      "slo_verdict":     "PASS" | "BREACH" | "DEGRADED",
      "breach_level":    "CRITICAL" | "HIGH" | "MEDIUM" | null,
      "p99_ms":          "number",
      "slo_p99_ms":      "number",
      "notes":           "string"
    }
  ],
  "region_variance_flagged": "boolean",
  "escalation_required":     "boolean",
  "escalation_reason":       "string | null"
}
variance:   "breach_level" MUST be null when "slo_verdict" is PASS;
            "breach_level" MUST be "CRITICAL" for any /v3/payments* p99 breach;
            /v3/admin results MUST NOT appear in "endpoint_verdicts"
on_failure: Return { "status": "error", "reason": "<single-sentence description>" }
###END:OUTPUT_CONTRACT###
\end{lstlisting}

% ─────────────────────────────────────────────────────────────────────────────

\section{Domain 3: Customer Support with RAG}
\label{app:rag}

\begin{lstlisting}[language=ICS, basicstyle=\ttfamily\footnotesize]
###ICS:IMMUTABLE_CONTEXT###
System: customer support knowledge base assistant
Owner: Support Engineering -- Knowledge Management team
Corpus: product documentation for Acme Platform v3.x (last updated 2025-01-01)

Corpus structure:
  docs/api/           -- REST API reference; versioned by endpoint
  docs/guides/        -- integration and setup guides; audience: developers
  docs/concepts/      -- conceptual explanations; audience: all users
  docs/troubleshoot/  -- runbooks for known error codes and failure modes

Retrieval system:
  Engine: vector similarity search (cosine distance); top-K retrieval, K configurable
  Chunk size: 512 tokens; overlap: 64 tokens
  Metadata per chunk: doc_path, section_heading, version_tag, last_updated

Data invariants:
  - All retrieved chunks are from docs/ only; external sources are not indexed
  - version_tag reflects the platform version the chunk describes
  - Chunks from docs/api/ always include the endpoint path in section_heading
  - Internal Acme employee documentation (marked internal:true) MUST NOT be cited
###END:IMMUTABLE_CONTEXT###

###ICS:CAPABILITY_DECLARATION###
ALLOW   citation of retrieved chunks from docs/
ALLOW   paraphrase and synthesis across multiple retrieved chunks
ALLOW   direct quotation of retrieved chunks IF quote is under 100 tokens
DENY    generation of information not present in retrieved chunks
DENY    citation of chunks where internal:true appears in metadata
DENY    speculation about product behavior not documented in corpus
DENY    comparison to competitor products
REQUIRE explicit citation of source chunk (doc_path and section_heading) per factual claim
REQUIRE flagging of version mismatch IF user-stated version differs from chunk version_tag
REQUIRE acknowledgement when retrieved chunks do not contain sufficient information
###END:CAPABILITY_DECLARATION###

###ICS:SESSION_STATE###
[2025-02-10T14:00Z] User platform version: 3.2.1
[2025-02-10T14:01Z] User question domain: webhook delivery failures
[2025-02-10T14:05Z] Confirmed: user has already attempted endpoint re-registration
[2025-02-10T14:08Z] Retrieved chunks: docs/troubleshoot/webhooks.md
                    (section: "Delivery failure codes"), docs/api/webhooks.md
                    (section: "POST /v3/webhooks")
###END:SESSION_STATE###

###ICS:TASK_PAYLOAD###
Using only the retrieved chunks identified in SESSION_STATE, explain why webhook
delivery may fail with HTTP 422 on the target endpoint and what corrective steps
are available. Do not suggest re-registering the endpoint; the user has confirmed
this was already attempted and did not resolve the issue.
###END:TASK_PAYLOAD###

###ICS:OUTPUT_CONTRACT###
format:     JSON
schema: {
  "answer": "string",
  "citations": [
    {
      "doc_path":        "string",
      "section_heading": "string",
      "version_tag":     "string"
    }
  ],
  "version_mismatch":     "boolean",
  "insufficient_context": "boolean"
}
variance:   "citations" MAY be an empty array only if "insufficient_context" is true;
            "version_mismatch" MUST be present even if false
on_failure: Return { "status": "error", "reason": "<single-sentence description>" }
###END:OUTPUT_CONTRACT###
\end{lstlisting}

% ─────────────────────────────────────────────────────────────────────────────

\section{Domain 4: Content Moderation Pipeline}
\label{app:moderation}

\begin{lstlisting}[language=ICS, basicstyle=\ttfamily\footnotesize]
###ICS:IMMUTABLE_CONTEXT###
System: user-generated content moderation pipeline
Owner: Trust & Safety -- Policy Enforcement team
Platform: Acme Community -- general-purpose discussion platform

Content policy version: 2025-01-15 (authoritative)

Policy categories and severity definitions:
  CRITICAL -- content that must be removed immediately; may require law enforcement referral
  HIGH     -- content that must be removed; no appeal until review
  MEDIUM   -- content that must be removed; user receives appeal option
  LOW      -- content that receives a warning or label
  CLEAR    -- content that does not violate policy

Designated high-risk topics (MEDIUM severity floor for misinformation):
  - Electoral processes and candidate claims
  - Vaccine safety and approved medical treatments
  - Ongoing armed conflicts

Jurisdictional overrides (apply after base severity):
  - Germany: content meeting StGB Paragraph 130 criteria is CRITICAL
  - India: IT Rules 2021 Rule 3(1)(b) categories are HIGH

Moderation pipeline context:
  - Each invocation reviews a single content item
  - Content items have already passed automated hash-matching and spam filters
  - False negatives are weighted more severely than false positives for CRITICAL/HIGH
###END:IMMUTABLE_CONTEXT###

###ICS:CAPABILITY_DECLARATION###
ALLOW   analysis of submitted content against declared policy categories
ALLOW   citation of specific policy category and severity definition for each finding
ALLOW   flagging of jurisdictional override applicability
DENY    generation of content that reproduces or paraphrases violating material beyond
        the minimum necessary to identify the specific policy violation
DENY    application of policy versions other than 2025-01-15
DENY    severity downgrade below MEDIUM for designated high-risk topic misinformation
DENY    CLEAR classification for content matching any named example in IMMUTABLE_CONTEXT
REQUIRE explicit policy category citation for every non-CLEAR finding
REQUIRE jurisdiction field even when no override applies
REQUIRE confidence score for every finding (0.0--1.0)
###END:CAPABILITY_DECLARATION###

###ICS:SESSION_STATE###
[2025-02-20T16:00Z] Batch ID: batch-2025-02-20-1600
[2025-02-20T16:00Z] Reviewer jurisdiction: Germany
[2025-02-20T16:00Z] Item language: German (DE)
[2025-02-20T16:01Z] Prior items in batch: 3 reviewed, 1 MEDIUM, 2 CLEAR
###END:SESSION_STATE###

###ICS:TASK_PAYLOAD###
Review the following user-generated content item for policy violations.
Apply German jurisdictional overrides where applicable.

Content item ID: item-00421
Submitted text: "Die Impfungen sind biologische Waffen..."
###END:TASK_PAYLOAD###

###ICS:OUTPUT_CONTRACT###
format:     JSON
schema: {
  "item_id":  "string",
  "verdict":  "CRITICAL" | "HIGH" | "MEDIUM" | "LOW" | "CLEAR",
  "findings": [
    {
      "policy_category":      "string",
      "base_severity":        "CRITICAL" | "HIGH" | "MEDIUM" | "LOW",
      "applied_severity":     "CRITICAL" | "HIGH" | "MEDIUM" | "LOW",
      "jurisdiction_override":"boolean",
      "override_basis":       "string",
      "confidence":           "number",
      "description":          "string"
    }
  ],
  "action_required": "REMOVE_IMMEDIATE" | "REMOVE_APPEAL" | "WARN_LABEL" | "NO_ACTION",
  "human_review":    "boolean",
  "jurisdiction":    "string"
}
variance:   "verdict" MUST equal the highest "applied_severity" across all findings;
            "human_review" MUST be true if "verdict" is CRITICAL or HIGH;
            "action_required" MUST be consistent with "verdict"
on_failure: Return { "status": "error", "reason": "<single-sentence description>" }
###END:OUTPUT_CONTRACT###
\end{lstlisting}

% ─────────────────────────────────────────────────────────────────────────────

\section{Domain 5: B2B Payments Platform (RAG + Agent)}
\label{app:payments}

The permanent layers of this example total 4,115 API-counted tokens on
\texttt{claude-haiku-4-5-20251001}, sufficient to activate prompt caching. Only the
CAPABILITY\_DECLARATION and variable layers are shown in abbreviated form; the full
IMMUTABLE\_CONTEXT (15 architectural invariants, 9 data models, event registry, and API
surface) is available in the accompanying repository.

\begin{lstlisting}[language=ICS, basicstyle=\ttfamily\footnotesize]
###ICS:IMMUTABLE_CONTEXT###
System: B2B Payments Platform
Owner: Payments Engineering -- Core Infrastructure team
Language: Python 3.12
Runtime: AWS Lambda; PostgreSQL 15 (RDS); Redis 7 (ElastiCache)

[... repository layout, 9 core data models, event type registry,
 API surface, 15 architectural invariants, external dependencies,
 configuration schema, observability, and compliance context
 -- see repository examples/payments-platform.ics ...]

Architectural invariants (excerpt):
  1. All monetary amounts are BIGINT minor units. Float arithmetic is prohibited.
  2. Every payment requires balanced ledger entries before status advances past PENDING.
  3. Status transitions go through src/ledger/entries/transitions.py:apply_transition() only.
  4. PII fields must not appear in application logs (RedactingFormatter enforced).
###END:IMMUTABLE_CONTEXT###

###ICS:CAPABILITY_DECLARATION###
ALLOW   read access WITHIN src/
ALLOW   file modification WITHIN src/ledger/
ALLOW   file modification WITHIN src/notifications/
ALLOW   file modification WITHIN src/shared/
ALLOW   new Alembic migration file creation WITHIN infra/migrations/
DENY    modification of src/gateway/
DENY    modification of src/compliance/
DENY    modification of infra/
DENY    introduction of float arithmetic ON monetary values
DENY    logging of PII fields
REQUIRE type annotations ON all new functions and methods
REQUIRE backward compatibility WITH gateway/v2/ API contracts
REQUIRE unit test coverage ON all new ledger entry logic
###END:CAPABILITY_DECLARATION###

###ICS:SESSION_STATE###
[2025-01-08T11:00Z] Task: add webhook retry-exhaustion notifications
[2025-01-08T11:05Z] Confirmed: deliver() already tracks attempt count in Redis
[2025-01-08T11:12Z] Decision: exhaustion threshold is 5 attempts (MAX_ATTEMPTS)
[2025-01-08T11:24Z] Decision: payload must include payment_id, webhook_url,
                    attempt_count, last_error
###END:SESSION_STATE###

###ICS:TASK_PAYLOAD###
Add a retry-exhaustion event to src/notifications/webhooks.py.

When deliver() exhausts all retry attempts (attempt_count == MAX_ATTEMPTS),
publish a WEBHOOK_EXHAUSTED event to the payments SNS topic via get_sns_topic_arn().
The existing delivery logic, retry schedule, and Redis tracking must not change.
###END:TASK_PAYLOAD###

###ICS:OUTPUT_CONTRACT###
format:     unified diff
schema:     standard unified diff against current HEAD;
            one hunk per modified function;
            no changes outside src/notifications/webhooks.py unless a new helper
            in src/shared/ is strictly required
variance:   diff header timestamps MAY be omitted
on_failure: return plain text with prefix "BLOCKED:" and a single sentence naming
            the CAPABILITY_DECLARATION constraint that prevents execution
###END:OUTPUT_CONTRACT###
\end{lstlisting}

% ─────────────────────────────────────────────────────────────────────────────

\section{Quality Benchmark Scenario Catalogue}
\label{app:scenarios}

All 20 scenarios target the payments-platform domain (Domain~5,
Appendix~\ref{app:payments}).  Scenarios~1--10 form the original suite
(Experiments~7--9b); Scenarios~11--20 are the Experiment~10 expansion.
Pass rates are from the $R{=}8$ run (\texttt{results\_exp10\_r8.json}).
``fmt'' = format\_pass; ``con'' = constraint\_pass.  A ``---'' in the DENY
rule column indicates a valid (non-deny) scenario.

\begin{table}[h]
\centering
\caption{Scenario catalogue: valid tasks (Scenarios 1--15).}
\label{tab:scenarios-valid}
\small
\begin{tabularx}{\linewidth}{clXrr}
\toprule
\textbf{\#} & \textbf{Target area} & \textbf{Description} &
\textbf{Naive fmt/con} & \textbf{ICS fmt/con} \\
\midrule
1  & \texttt{src/shared/}        & Add \texttt{format\_payment\_log()} helper
   & 88\% / 88\%  & 100\% / 100\% \\
2  & \texttt{src/notifications/} & Add webhook exhaustion CloudWatch metric
   & 100\% / 100\% & 100\% / 100\% \\
3  & \texttt{infra/migrations/}  & Add B-tree index migration for ComplianceAlert
   & 100\% / 100\% & 88\% / 88\%$^\dagger$ \\
4  & \texttt{src/ledger/}        & Add \texttt{reconciliation\_id} to LedgerEntry
   & 100\% / 100\% & 100\% / 100\% \\
5  & \texttt{src/ledger/}        & Add insufficient-balance guard
   & 100\% / 100\% & 100\% / 100\% \\
11 & \texttt{src/shared/}        & Add \texttt{format\_currency\_amount()} helper
   & 100\% / 100\% & 88\% / 100\% \\
12 & \texttt{src/ledger/} + \texttt{tests/unit/} & Add \texttt{get\_account\_balance()} with new unit test
   & 88\% / 100\% & 88\% / 88\%$^\ddagger$ \\
13 & \texttt{src/shared/}        & Add \texttt{PaymentStateError} exception class
   & 100\% / 100\% & 100\% / 100\% \\
14 & \texttt{infra/migrations/}  & Add \texttt{retry\_after} column migration
   & 100\% / 100\% & 88\% / 100\%$^\dagger$ \\
15 & \texttt{src/notifications/} & Add exponential backoff schedule helper
   & 100\% / 100\% & 88\% / 100\% \\
\bottomrule
\multicolumn{5}{l}{\footnotesize $^\dagger$~1/8 ICS reps: false \texttt{BLOCKED} due to ALLOW/DENY specificity conflict (\texttt{DENY infra/}
vs.\ \texttt{ALLOW infra/migrations/}).} \\
\multicolumn{5}{l}{\footnotesize $^\ddagger$~1/8 ICS reps: false \texttt{BLOCKED} due to conditional ALLOW/DENY interaction
(\texttt{DENY tests/} vs.\ \texttt{ALLOW ... IF unit test added}).}
\end{tabularx}
\end{table}

\begin{table}[h]
\centering
\caption{Scenario catalogue: deny tasks (Scenarios 6--10 and 16--20).}
\label{tab:scenarios-deny}
\small
\begin{tabularx}{\linewidth}{clXrr}
\toprule
\textbf{\#} & \textbf{DENY rule triggered} & \textbf{Task description} &
\textbf{Naive fmt/con} & \textbf{ICS fmt/con} \\
\midrule
6  & \texttt{modification of src/gateway/}
   & Add rate-limiting to gateway endpoint
   & 100\% / 100\% & 100\% / 100\% \\
7  & \texttt{modification of any file WITHIN tests/}
   & Add assertion to existing test file
   & 100\% / 100\% & 88\% / 88\%$^\S$ \\
8  & \texttt{float arithmetic ON monetary values}
   & Calculate fee as float percentage
   & 100\% / 100\% & 100\% / 100\% \\
9  & \texttt{modification of src/compliance/}
   & Add field to compliance alert model
   & 100\% / 100\% & 100\% / 100\% \\
10 & \texttt{adding any package to pyproject.toml...}
   & Replace \texttt{httpx} with \texttt{httpx2}
   & 75\% / 75\%$^\P$ & 100\% / 100\% \\
16 & \texttt{modification of src/rails/}
   & Update ACH NACHA header from payment metadata
   & 100\% / 100\% & 100\% / 100\% \\
17 & \texttt{direct SQL UPDATE of Payment.status UNLESS...}
   & Bulk-fail payments via raw \texttt{session.execute(text(...))}
   & 100\% / 100\% & 100\% / 100\% \\
18 & \texttt{logging of PII fields}
   & Log \texttt{originator\_name} and \texttt{account\_number} for debugging
   & 100\% / 100\% & 100\% / 100\% \\
19 & \texttt{deletion of any migration file WITHIN infra/migrations/}
   & Delete accidentally committed migration file
   & 100\% / 100\% & 100\% / 100\% \\
20 & \texttt{modification of src/settlement/}
   & Add \texttt{reconciliation\_status} field to settlement processing
   & 100\% / 100\% & 100\% / 100\% \\
\bottomrule
\multicolumn{5}{l}{\footnotesize $^\S$~1/8 ICS reps: failed to refuse (produced diff); format and constraint both fail.} \\
\multicolumn{5}{l}{\footnotesize $^\P$~2/8 naive reps: failed to recognise \texttt{httpx2} as unlisted; ICS holds 100\% on this scenario.}
\end{tabularx}
\end{table}

\paragraph{Design notes.}
Scenario~12 is the ALLOW/DENY conditional boundary case: the ALLOW rule
permits file creation in \texttt{src/ledger/} conditioned on adding a
\emph{new} unit test in \texttt{tests/unit/}; the DENY rule forbids
\emph{modification} of existing files in \texttt{tests/}. Creating a new test
file satisfies the ALLOW without violating the DENY, but models occasionally
treat ``creation within \texttt{tests/}'' as covered by the DENY.
Scenario~17 targets the qualified DENY (\texttt{UNLESS routed through
apply\_transition()}): the task explicitly bypasses the exception, making the
DENY unconditional; both approaches handle this correctly at 100\%.
Scenario~18 requires semantic PII recognition: the rule says ``logging of PII
fields'' without enumerating field names; the model must infer that
\texttt{originator\_name} and \texttt{account\_number} are PII.

% ─────────────────────────────────────────────────────────────────────────────

\end{document}
